Lossy compression of DNA sequencing quality data

### 2015-05-13 manuscript/q-compression.tex
Added eps

\begin{figure}[!tpb]
\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
\caption{Quality profiles obtained by $k$-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using $k$ = 128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. One can also see sporadic low-position values in a handful
  of profiles, likely capturing intermittent problems in the
  sequencing process affecting thousands of reads at a
  time.}\label{fig:profiles_128}
\end{figure}

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
  denote the corresponding lossless compression using \textsc{bz}ip2,
  with the black arrow corresponding to original data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call in the
\textit{Mus musculus} data set, and requires more storage than the
original \textsc{ascii} quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per read. The read length of the \textit{Mus musculus} data set is
only 26, so the 7th-degree polynomial regression is storing six more
bytes than necessary for lossless encoding of the quality data.

@@

\begin{figure}[!tbp]
\centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. Reads were trimmed using Sickle. The total
  number of bases filtered by each compression method is compared with
  the number of bases filtered using the uncompressed quality values.}
  \label{fig:preprocessing}
\end{figure}

@@

\begin{figure}[!tbp]
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}

>>>>

\begin{figure}[!tpb]
\centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
\caption{Quality profiles obtained by $k$-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using $k$ = 128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. One can also see sporadic low-position values in a handful
  of profiles, likely capturing intermittent problems in the
  sequencing process affecting thousands of reads at a
  time.}\label{fig:profiles_128}
\end{figure}

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.eps}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
  denote the corresponding lossless compression using \textsc{bz}ip2,
  with the black arrow corresponding to original data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call in the
\textit{Mus musculus} data set, and requires more storage than the
original \textsc{ascii} quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per read. The read length of the \textit{Mus musculus} data set is
only 36 nt, so the 7th-degree polynomial regression is storing six more
bytes than necessary for lossless encoding of the quality data.

@@

\begin{figure}[!tbp]
\centerline{\includegraphics[width=3.65in]{preprocessing_results.eps}}
\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. Reads were trimmed using Sickle. The total
  number of bases filtered by each compression method is compared with
  the number of bases filtered using the uncompressed quality values.}
  \label{fig:preprocessing}
\end{figure}

@@

\begin{figure}[!tbp]
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.eps}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}
